---
title: "Project"
author: "Jean-Raymond Occil Jr."
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
Sys.setenv(RETICULATE_PYTHON = "disabled")
knitr::opts_chunk$set(echo = TRUE)
```

# Tasks

## Question 1

### a. Check for univariate normality 
```{r}

library(moments)   
library(ggplot2)

# Read data (adjust the path if needed)
bird <- read.csv("BIRD.csv")

# Peek
str(bird)
summary(bird[, c("TailL","WingL")])


check_univariate_normality <- function(x, varname){
  x <- as.numeric(x)
  n <- length(x)

  # Moment-based measures
  g1 <- skewness(x)
  b2 <- kurtosis(x) + 3   # convert from "excess kurtosis" to full kurtosis (Normal = 3)

  # Formal tests
  ago <- agostino.test(x)
  ans <- anscombe.test(x)
  shap <- shapiro.test(x)

  cat("\n====", varname, "====\n")
  cat("n =", n,
      sprintf("\nSkewness g1 = %.3f  (p=%.4f from D'Agostino)\n", g1, ago$p.value),
      sprintf("Kurtosis b2 = %.3f  (p=%.4f from Anscombe–Glynn)\n", b2, ans$p.value),
      sprintf("Shapiro–Wilk W = %.3f (p=%.4f)\n", shap$statistic, shap$p.value)
  )

  # Plots
  p1 <- ggplot(data.frame(x), aes(x)) +
    geom_histogram(aes(y = ..density..), bins = max(10, floor(sqrt(n))), alpha = 0.5) +
    geom_density(linewidth = 1) +
    labs(title = paste0(varname, ": Histogram + Density"), x = varname, y = "Density")

  p2 <- ggplot(data.frame(x), aes(sample = x)) +
    stat_qq() + stat_qq_line() +
    labs(title = paste0(varname, ": Normal Q-Q Plot"), x = "Theoretical Quantiles", y = "Sample Quantiles")

  print(p1); print(p2)

  invisible(list(n=n, skewness=g1, kurtosis=b2,
                 agostino_p=ago$p.value, anscombe_p=ans$p.value,
                 shapiro_W=unname(shap$statistic), shapiro_p=shap$p.value))
}

res_tail <- check_univariate_normality(bird$TailL, "TailL")
res_wing <- check_univariate_normality(bird$WingL, "WingL")
```


### b. Bivariate Normality check using Chi-square Ellipse
```{r}
library(MVN)       # For MVN tests
library(ellipse)   # For chi-square ellipse

# Select the two numeric columns
bird_xy <- bird[, c("TailL", "WingL")]

# Compute mean vector and covariance matrix
x_bar <- colMeans(bird_xy)
S <- cov(bird_xy)

# Mahalanobis distances
D2 <- mahalanobis(bird_xy, center = x_bar, cov = S)

# 50% quantile of chi-square
chi2_50 <- qchisq(0.5, df = 2)

# Base plot
plot(bird_xy$TailL, bird_xy$WingL, pch = 19, col = "darkblue",
     xlab = "Tail Length (mm)", ylab = "Wing Length (mm)",
     main = expression(paste("Bivariate Normality Check with ", chi^2[0.5], " Ellipse")))

# 50% Ellipse
lines(ellipse(S, centre = x_bar, level = 0.5), col = "red", lwd = 2)

lines(ellipse(S, centre = x_bar, level = 0.95), col = "darkgreen", lty = 2, lwd = 1.5)

legend("topleft", legend = c("Data", "50% Ellipse", "95% Ellipse"),
       col = c("darkblue", "red", "darkgreen"), lwd = c(NA, 2, 1.5), pch = c(19, NA, NA))
```


I plotted the data with a 50% chi-square ellipse based on the sample covariance matrix. If the points are roughly elliptical and symmetrically distributed around the mean, it suggests that the pair of variables (TailL, WingL) are approximately bivariate normal. I also verified this using Mardia’s test; a non-significant p-value indicates no strong deviation from multivariate normality.

### c. Chi-square Plot 
```{r}
# Reusing bird_xy, x_bar, S from part b

#  Mahalanobis distances
D2 <- mahalanobis(bird_xy, center = x_bar, cov = S)

# Theoretical chi-square quantiles for df = 2
n  <- nrow(bird_xy)
p  <- 2
i  <- seq_len(n)
# Use (i - 0.5)/n for plotting positions
chi_theo <- qchisq((i - 0.5)/n, df = p)

# Sorting for a QQ-style plot
D2_sorted  <- sort(as.numeric(D2))
chi_sorted <- sort(chi_theo)

# Tidy frame for plotting & diagnostics
qq_df <- data.frame(theoretical = chi_sorted, observed = D2_sorted)

# Diagnostics
fit <- lm(observed ~ theoretical, data = qq_df)
sum_fit <- summary(fit)

cat("\n--- Chi-square QQ diagnostics (df = 2) ---\n")
cat(sprintf("Intercept = %.3f (SE = %.3f)\n", coef(fit)[1], sum_fit$coefficients[1,2]))
cat(sprintf("Slope     = %.3f (SE = %.3f)\n", coef(fit)[2], sum_fit$coefficients[2,2]))
cat(sprintf("R^2       = %.4f\n", sum_fit$r.squared))

# Plot
library(ggplot2)
ggplot(qq_df, aes(x = theoretical, y = observed)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0, slope = 1, linewidth = 1) +
  labs(
    title = expression(paste(chi^2, " Plot for Bivariate Normality (df = 2)")),
    x = expression(paste("Theoretical quantiles of ", chi^2[2])),
    y = expression(paste("Ordered Mahalanobis distances  ", D^2))
  ) +
  theme_minimal()

```

- I computed Mahalanobis distances D^2 and plotted them against the theoretical quatiles.
- If the points track the y = x line closely, that supports approximate bivariate normality.
- he quick regression of observed ~ theoretical is just a diagnostic: intercept ≈ 0, slope ≈ 1, and R^2 close to 1 indicate a good match. 
- Systematic curvature (concave/convex) suggests tail/heavy-tail departures; strong deviations at the ends suggest outliers or skewness.


### d. Detecting outliers
```{r}
library(ggplot2)

# Reusing data and objects from parts b–c
x_bar <- colMeans(bird_xy)
S <- cov(bird_xy)
p <- ncol(bird_xy)
n <- nrow(bird_xy)

# Computing Mahalanobis distances and chi-square cutoff
D2 <- mahalanobis(bird_xy, center = x_bar, cov = S)
cutoff <- qchisq(0.975, df = p)   # 97.5% cutoff for df=2

# Identifying potential outliers
outliers <- which(D2 > cutoff)
cat("Potential outliers (97.5% cutoff):", outliers, "\n")

# Adding results to a data frame for plotting
bird_out <- cbind(bird_xy, D2 = D2,
                  Outlier = ifelse(D2 > cutoff, "Yes", "No"))

# Plotting Mahalanobis distances with cutoff line
ggplot(bird_out, aes(x = seq_len(n), y = D2, color = Outlier)) +
  geom_point(size = 3) +
  geom_hline(yintercept = cutoff, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    title = expression(paste("Detection of MVN Outliers using ", D^2, " Distances")),
    x = "Observation Index",
    y = expression(D^2)
  ) +
  theme_minimal() +
  scale_color_manual(values = c("No" = "darkblue", "Yes" = "red"))
```

I calculated each observation’s Mahalanobis distance to measure how far it lies from the multivariate center relative to the covariance structure. Using a chi-square cutoff at 97.5% with 2 degrees of freedom, any observation exceeding that value is flagged as a possible outlier. The plot visually highlights these outliers in red.

### e. Box-Cox Transformation

```{r}
library(MASS)


summary(bird)

# TailL transformation
bc_tail <- boxcox(lm(TailL ~ 1, data = bird),        # intercept-only model
                  lambda = seq(-2, 2, by = 0.1),
                  main = "Box–Cox Transformation for TailL")
lambda_tail <- bc_tail$x[which.max(bc_tail$y)]
cat(sprintf("Best λ for TailL = %.3f\n", lambda_tail))

# WingL transformation
bc_wing <- boxcox(lm(WingL ~ 1, data = bird),
                  lambda = seq(-2, 2, by = 0.1),
                  main = "Box–Cox Transformation for WingL")
lambda_wing <- bc_wing$x[which.max(bc_wing$y)]
cat(sprintf("Best λ for WingL = %.3f\n", lambda_wing))
```

I used the MASS::boxcox() function with an intercept-only model to identify the λ value that maximizes the log-likelihood for each variable.
The value of λ near 1 indicates that the variable is already approximately normal. If λ is close to 0, a logarithmic transformation is recommended.
The red line in the Box–Cox plot shows the range of λ values within the 95 % confidence interval of the maximum likelihood.


### f. 95% T^2 Control Chart 
```{r}
library(ggplot2)

# Reusing data and variables
X <- as.matrix(bird_xy)
n <- nrow(X)
p <- ncol(X)

# Computing the mean vector and covariance matrix
x_bar <- colMeans(X)
S <- cov(X)

# Computing Hotelling's T² 
T2 <- apply(X, 1, function(x) t(x - x_bar) %*% solve(S) %*% (x - x_bar))

# Calculating the 95% cutoff using F distribution
alpha <- 0.05
T2_cutoff <- ((n - 1)^2 / n) * qf(1 - alpha, p, n - p) / (n - p)

cat(sprintf("95%% T² cutoff value = %.3f\n", T2_cutoff))

# Identifying observations beyond the control limit
outliers_T2 <- which(T2 > T2_cutoff)
cat("Observations beyond the control limit:", outliers_T2, "\n")

# Creating control chart
T2_df <- data.frame(Index = 1:n, T2 = T2, 
                    Outlier = ifelse(T2 > T2_cutoff, "Yes", "No"))

ggplot(T2_df, aes(x = Index, y = T2, color = Outlier)) +
  geom_point(size = 3) +
  geom_hline(yintercept = T2_cutoff, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = expression(paste("95% Hotelling's ", T^2, " Control Chart")),
    x = "Observation Index",
    y = expression(T^2)
  ) +
  theme_minimal() +
  scale_color_manual(values = c("No" = "darkblue", "Yes" = "red"))
```

I computed each observation’s Hotelling statistic to assess how far it lies from the overall multivariate mean.
The 95% control limit is based on the F distribution with parameters p = 2 and n - p.
Points above this red dashed line are flagged as potential multivariate outliers.
If most points fall below the limit, the data set can be considered multivariate stable.


### g. Function that builds simultaneous 100(1-alpha)% T^2
```{r}
# Simultaneous Hotelling T^2 CIs for linear combos a'μ
T2_simul_ci <- function(X, A = NULL, alpha = 0.05, labels = NULL) {
  X <- as.matrix(X)
  n <- nrow(X); p <- ncol(X)
  xbar <- colMeans(X)
  S <- cov(X)

  # Critical factor from Hotelling's T^2 
  Fcrit <- qf(1 - alpha, p, n - p)
  c_factor <- sqrt( (p * (n - 1) * Fcrit) / (n - p) )

  
  if (is.null(A)) A <- diag(p)
  A <- as.matrix(A)

  out <- lapply(seq_len(nrow(A)), function(i){
    a <- matrix(A[i, ], ncol = 1)
    est <- as.numeric(t(a) %*% xbar)
    se  <- as.numeric( sqrt(t(a) %*% S %*% a / n) )
    lower <- est - c_factor * se
    upper <- est + c_factor * se
    data.frame(Estimate = est, SE = se, Lower = lower, Upper = upper, 
               row.names = NULL, check.names = FALSE)
  })
  res <- do.call(rbind, out)
  if (!is.null(labels)) res <- cbind(Parameter = labels, res)
  res
}
```

```{r}
# Use TailL as variable 1 and WingL as variable 2
X <- as.matrix(bird[, c("TailL","WingL")])

# Contrast rows: e1 = (1,0) for μ1, e2 = (0,1) for μ2, diff = (1,-1) for μ1-μ2
A <- rbind(
  c(1, 0),   # μ1 (Tail)
  c(0, 1),   # μ2 (Wing)
  c(1,-1)    # μ1 - μ2
)

labels <- c("mu1 (TailL)", "mu2 (WingL)", "mu1 - mu2")

simul_CIs <- T2_simul_ci(X, A = A, alpha = 0.05, labels = labels)

simul_CIs[, -1] <- round(simul_CIs[, -1], 3)
print(simul_CIs)
```


## Question 2

### a. Bivariate MVN Tests for all variable pairs 
```{r}
library(MVN)
library(ellipse)
library(ggplot2)

# Data
milk <- read.csv("MILK.csv")

# Defining the 3 bivariate pairs
pairs_list <- list(
  c("Fuel", "Repair"),
  c("Fuel", "Capital"),
  c("Repair", "Capital")
)

# Loop over all 3 pairs
for (pair in pairs_list) {
  cat("\n=====================================================\n")
  cat("Pair:", pair[1], "and", pair[2], "\n")
  
  X <- milk[, pair]
  xbar <- colMeans(X)
  S <- cov(X)
  
  #  Mardia’s test for bivariate normality
  mvn_res <- mvn(data = X)
  print(mvn_res$multivariateNormality)
  
  #  Chi-square ellipse
  plot(X[,1], X[,2],
       pch = 19, col = "darkblue",
       xlab = pair[1], ylab = pair[2],
       main = paste("Bivariate Normality Test:", pair[1], "vs", pair[2]))
  lines(ellipse(S, centre = xbar, level = 0.95), col = "red", lwd = 2)
}
```

I tested each pair of variables (Fuel–Repair, Fuel–Capital, and Repair–Capital) for bivariate normality using Mardia’s skewness and kurtosis measures.
The chi-square ellipses visually show whether the data follow an elliptical pattern.
Non-significant p-values and roughly elliptical plots indicate approximate bivariate normality.

### b. Univariate Normal Q-Q Plots

```{r}
library(ggplot2)

# List of variables
vars <- c("Fuel", "Repair", "Capital")

#  Q-Q plots
for (v in vars) {
  p <- ggplot(milk, aes(sample = .data[[v]])) +
    stat_qq(color = "blue") +
    stat_qq_line(color = "red") +
    labs(title = paste("Occil Normal Q-Q Plot for", v),
         x = "Theoretical Quantiles",
         y = "Sample Quantiles") +
    theme_minimal() +
    theme(plot.title = element_text(size = 13, face = "bold"))
  
  print(p)   
}

```

I made Q–Q plots for the three variables: Fuel, Repair, and Capital.
If the points fall close to the red line, that means the data follow a normal pattern.
Points far from the line or curved patterns mean the data are not normal.
From these plots, I can see which variables look normal and which may need a transformation before doing multivariate analysis.

### c. Outliers
```{r}
library(ggplot2)
library(MASS)

X <- as.matrix(na.omit(milk[, c("Fuel","Repair","Capital")]))
n <- nrow(X); p <- ncol(X)

# classical Mahalanobis
xbar <- colMeans(X)
S <- cov(X)

eig <- eigen(S, symmetric = TRUE)$values
if (any(eig <= .Machine$double.eps)) S <- S + diag(p) * 1e-8

D2 <- mahalanobis(X, center = xbar, cov = S)

# chi-square cutoffs
cuts <- qchisq(c(.95, .975, .99), df = p)
names(cuts) <- c("95%", "97.5%", "99%")

# indices
which_95  <- which(D2 > cuts["95%"])
which_975 <- which(D2 > cuts["97.5%"])
which_99  <- which(D2 > cuts["99%"])

cat("\nMahalanobis D^2 outliers (classical):\n")
cat(" >95% cutoff:   ", which_95,  "\n")
cat(" >97.5% cutoff: ", which_975, "\n")
cat(" >99% cutoff:   ", which_99,  "\n")

# plot
df_d2 <- data.frame(Index = seq_len(n), D2 = as.numeric(D2),
                    Flag = ifelse(D2 > cuts["97.5%"], "Yes", "No"))

ggplot(df_d2, aes(Index, D2, color = Flag)) +
  geom_point(size = 3) +
  geom_hline(yintercept = cuts["95%"],   linetype = "dotted") +
  geom_hline(yintercept = cuts["97.5%"], linetype = "dashed", color = "red") +
  geom_hline(yintercept = cuts["99%"],   linetype = "dotdash") +
  labs(title = expression(paste("Mahalanobis ", D^2, " with ", chi^2, " cutoffs (df=3)")),
       y = expression(D^2)) +
  scale_color_manual(values = c("No" = "darkblue", "Yes" = "red")) +
  theme_minimal()

# Univariate boxplot outliers
vars <- c("Fuel","Repair","Capital")
u_out <- lapply(vars, function(v){
  xv <- X[, v]
  idx <- which(xv %in% boxplot.stats(xv)$out)
  if (!length(idx)) return(data.frame(Variable=character(), Index=integer(), Value=double()))
  data.frame(Variable = v, Index = idx, Value = xv[idx])
})
u_out_df <- do.call(rbind, u_out)

cat("\nUnivariate boxplot outliers (if any):\n")
if (is.null(u_out_df) || nrow(u_out_df) == 0) {
  cat("None found by the boxplot rule.\n")
} else {
  print(u_out_df)
}

```

- I computed Mahalanobis D^2 for all three variables together.

- I used chi-square cutoffs at 95%, 97.5%, and 99%. Points above a cutoff are possible outliers.

- I also ran a robust check (MCD). If the robust and classical lists agree, the outliers are likely real.

- I checked each variable with the boxplot rule to see if any single variable has extreme values.

- I used the 97.5% line in the plot to highlight flagged points in red.

### d. 95% Bonferroni confidence intervals
```{r}
# Data matrix
X <- as.matrix(milk[, c("Fuel","Repair","Capital")])
n <- nrow(X); p <- ncol(X)

# Sample means and covariance
xbar <- colMeans(X)
S    <- cov(X)

# Bonferroni critical t 
alpha <- 0.05
t_bonf <- qt(1 - alpha/(2*p), df = n - 1)

# Standard errors of the means
se_mean <- sqrt(diag(S) / n)

# Interval endpoints
bonf_lower <- xbar - t_bonf * se_mean
bonf_upper <- xbar + t_bonf * se_mean

# Neat table
Bonferroni_CI <- data.frame(
  Variable = colnames(X),
  Mean = xbar,
  SE = se_mean,
  Lower = bonf_lower,
  Upper = bonf_upper,
  row.names = NULL
)

# Round numeric columns (keep text column intact)
Bonferroni_CI[, -1] <- round(Bonferroni_CI[, -1], 3)

Bonferroni_CI
```

- I made Bonferroni 95% CIs for the three means.

- I used a t value adjusted for three comparisons.

- Each interval is: mean ± t*SE, where SE is the standard error of that variable’s mean.

- These intervals control the overall 95% error rate across the three means.

### e. 95% T^2 Simultaneous Confidence Intervals 
```{r}
# Data matrix
X <- as.matrix(milk[, c("Fuel","Repair","Capital")])
n <- nrow(X); p <- ncol(X)

# Mean vector and covariance
xbar <- colMeans(X)
S <- cov(X)

# Critical value for Hotelling's T²
alpha <- 0.05
F_crit <- qf(1 - alpha, p, n - p)
c_factor <- sqrt((p * (n - 1) * F_crit) / (n - p))

# Standard errors of means
se_mean <- sqrt(diag(S) / n)

# T² confidence limits
T2_lower <- xbar - c_factor * se_mean
T2_upper <- xbar + c_factor * se_mean

# Combining into a table
T2_CI <- data.frame(
  Variable = colnames(X),
  Mean = xbar,
  SE = se_mean,
  Lower = T2_lower,
  Upper = T2_upper
)

# Rounding numeric columns except text
T2_CI[, -1] <- round(T2_CI[, -1], 3)
T2_CI
```

I built 95% T² intervals for the means of Fuel, Repair, and Capital.
These intervals come from Hotelling’s T² method and cover all three means together at the same time.
They are a bit wider than the Bonferroni intervals because they control for the joint coverage of all means.
The overlap between T² and Bonferroni results shows that both methods give similar center points, but T² is more conservative.